Measuring the double-drift illusion with hand trajectories

### 2021-12-15 Manuscript.Rmd
edits from Patrick

```{r fig4, echo=FALSE, message=FALSE, fig.width=6, fig.height=6, fig.cap="\\label{fig:fig4}**Figure 4:** Re-tracing Data. **A-D:** Example trajectories and reset points. All trajectory samples in transparent light blue (denser samples result in darker color). Dark red lines are, pruned, interpolated and filtered trajectories (see Methods). Purple circles indicate reset points. **A:** Trajectory with several sudden resets. **B:** 'Hit-the-wall' reset. **C:** Slow return reset. **D:** Continued accumulation of the illusion without any reset. **E:** Distribution of all 827 detected reset points. **F:** Using the angle from the origin of the trace to a point along the trajectory at half the distance between origin and reset point as a measure of illusion strength, we can compare this data to other data. For this graph, we exclude one participant who produced very few resets. The illusion is weaker than in Experiment 1 ($K=0.60$, blue line) and its strength varies between participants, but it is nevertheless clearly there."}
#plotExampleData()
plotResetPoints()
```

@@

These first two models assumed that resets were triggered at fixed locations in space or time, but the next two propose that the resets are random processes that occur with some distribution across space or time. We used gamma distributions to capture these processes (also with 2 parameters: rate; the chance of an event per second, and shape; how many events are needed, in this case, until a reset is triggered) on the X coordinates and time points of resets as well (see Fig 5A, marginal distributions; Fig 5B,C). To compare across the four models, we then calculated the likelihoods of their predictions (based on the underlying distributions' probability density functions) using all resets together and also within individual participants. From these likelihood values, we calculated an AIC to determine a relative probability for each model.

@@

In our second experiment, we asked participants to draw the perceived path of the stimulus after observing a single pass. The orientation of the drawn traces reflect the strength of the illusion, and in a majority of trials (63.8%) there was a clearly indentifiable reset point, of either a "return" or a "hit-the-wall" type. We then set out to test if these spontaneous resets were triggered by a limiting distance of drift away from the physical location, or by a specific limit of time. Neither of these provided a viable explanation of the pattern of resets. Resets points are broadly distributed, suggesting that resets are not linked to the degree of illusory offset or time from stimulus onset, but instead may be caused by a more "random" process.

Recently, it has been suggested that resets can be triggered by a distraction of attention [@Nakayama2020]. The participants in their experiment also retraced the perceived path and reported “return” resets. Nakayama and Holcombe [-@Nakayama2020] claim that planned eye movements can trigger a “return” reset back to the physical path. In our experiment, we did not control for eye-movements, so we can't directly investigate whether eye movements accompanied the “return” resets that participants reported. However, the random distribution of resets throughout time are in line with a mechanism based on distracted attention, which may involve eye-movement planning and execution. The rate of resets that we found here, ~0.75 / _s_, are too infrequent for microsaccades (~3 / _s_) so the distractions may be more significant than these small eye movements. On the other hand, gamma distributions [e.g. @McGill1965; @Wolfe2002] do model a random occurence of events (rate parameter) where only some fraction of these trigger an action (shape parameter). This could be consistent with the more frequent microsaccades happening randomly at a faster rate where only some of there would trigger a reset. In other words, the gamma distribution is not only a good fit to our reset data but it may reveal the processes underlying the resets.

The fit showed that reset times were well modeled by a broad gamma distribution, with resets occurring randomly in time (set by the rate and shape parameter) with an average rate of 1 reset every 1.3 _s_ over the 3 to 4 _s_ duration of the trials. However, we have to qualify our conclusions given the noisy nature of hand-drawn trajectories as a measure of the perceived trajectory and the reset points. To gauge the generality of our results, we compared them to those from a similar study that analyzed trajectory endpoints [pre-print: @Liu2021] to determine the presence and size of spontaneous resets. In their study, the internal speed of the Gabor texture was adjusted across conditions and participants so that the illusion always had a local direction of 45° compared to its actual vertical direction. This then generated a highly predictable offset of the end of the trajectory that would be $\sqrt{2}$ of the path length if there were no spontaneous resets. Any offset less than that revealed the presence and size of spontaneous resets. Liu et al. [-@Liu2021] tested different durations and path lengths for the first segment of the Gabor’s motion (see Fig. 6A). Then a 350 _ms_ gap in stimulus presentation is introduced, which is known to evoke a full reset to the physical location [@Lisi2015]. The difference between the perceived points where the Gabor disappeared (illusion minus spontaneous resets) and reappeared (veridical), can be used to assess spontaneous resets.

The data from 10 participants showed that for a fixed path and illusion strength, longer durations produced smaller illusory offsets at the end (Fig 6B, round dots). This can be explained if resets occurred randomly over time, so that longer durations would have more resets and a smaller final offset at the endpoint. Liu et al. [-@Liu2021] modeled this with a Poisson process having a constant chance of reset of ~0.7 / _s_ (Fig 6B, light diamonds). In comparison, the gamma distribution fit to our data here (rate=$2.81$, shape=$3.74$), gave a ~0.75 / _s_ chance of a reset. In addition, we simply used the parameters of the gamma distribution fit to our data to generate the PSEs that should be found in Liu et al.'s experiment (Fig 6B, dark diamonds). We see that the two findings, based on very different approaches, are mostly in agreement with each other.

```{r fig6, echo=FALSE, message=FALSE, fig.width=6, fig.height=4, fig.cap="\\label{fig:fig6}**Figure 6:** Spontaneous resets without hand trajectories. Experiment and Results based on data from Liu et al. [-@Liu2021], re-analyzed here. **A:** By ensuring the perceived path of a double-drift stimulus is straight down (teal arrows), any gap-induced reset, should be equal to the accumulated illusion if there are no spontaneous resets (black arrow), but smaller if there are (light blue arrow). Gaps were presented after 1, 2 or 3 seconds, and 2 or 4 dva real paths, which allows psychophysics to quantify the amount of illusion left after several periods of observing a double-drift stimulus. **B:** The endpoints of the first segments indicate the size of the illusion and the difference between these values and the expected value (1.4 or 2.8 _dva_) indicates the size of any intervening spontaneous resets. This shows that the size of the illusion decreases over time when path length and local illusion angle are kept constant. These data are in line with resets occurring randomly over time at ~0.7 / _s_."}
plotJumpPSEs(target = 'inline', individualDots=FALSE, modelDots=FALSE, gammaDots=FALSE, modelLine=TRUE, gammaLine=TRUE)
```

>>>>

```{r fig4, echo=FALSE, message=FALSE, fig.width=6, fig.height=6, fig.cap="\\label{fig:fig4}**Figure 4:** Re-tracing Data. **A-D:** Example trajectories and reset points. All trajectory samples in transparent light blue (denser samples result in darker color). Dark red lines are, pruned, interpolated and filtered trajectories (see Methods). Purple circles indicate reset points. **A:** Trajectory with several sudden resets. **B:** 'Hit-the-wall' reset. **C:** Slow return reset. **D:** Continued accumulation of the illusion without any reset. **E:** Distribution of all 827 detected reset points. **F:** Using the angle from the origin of the trace to a point along the trajectory at half the distance between origin and reset point as a measure of illusion strength, we can compare these data to other data. For this graph, we exclude one participant who produced very few resets. The illusion is weaker than in Experiment 1 ($K=0.60$, blue line) and its strength varies between participants, but it is nevertheless clearly there."}
#plotExampleData()
plotResetPoints()
```

@@

These first two models assumed that resets were triggered at fixed locations in space or time, but the next two propose that the resets are random processes that occur with some distribution across space or time. We used gamma distributions to capture these processes (these have 2 parameters: rate, the chance of an event per second; and shape, how many events are needed, in this case, until a reset is triggered) on the X coordinates and time points of resets as well (see Fig 5A, marginal distributions; Fig 5B,C). To compare across the four models, we then calculated the likelihoods of their predictions (based on the underlying distributions' probability density functions) using all resets together and also within individual participants. From these likelihood values, we calculated an AIC to determine a relative probability for each model.

@@

In our second experiment, we asked participants to draw the perceived path of the stimulus after observing a single pass. The orientation of the drawn traces reflect the strength of the illusion, and in a majority of trials (63.8%) there was a clearly indentifiable reset point, of either a "return" or a "hit-the-wall" type. We then set out to test if these spontaneous resets were triggered by a limiting distance of drift away from the physical location, or by a specific limit of time. Neither of these provided a viable explanation of the pattern of resets. Resets points are broadly distributed, suggesting that resets are not linked to the degree of illusory offset or to the time from stimulus onset, but instead may be caused by a more "random" process.

Recently, it has been suggested that resets can be triggered by a distraction of attention [@Nakayama2020]. The participants in their experiment also retraced the perceived path and reported “return” resets. Nakayama and Holcombe [-@Nakayama2020] claim that planned eye movements can trigger a “return” reset back to the physical path. In our experiment, we did not control for eye-movements, so we can't directly investigate whether eye movements accompanied the “return” resets that participants reported. However, the random distribution of resets throughout time are in line with a mechanism based on distracted attention, which may involve eye-movement planning and execution. The rate of resets that we found here, ~0.75 / _s_, is too infrequent for microsaccades (~3 / _s_) so the distractions may be more significant than these small eye movements. On the other hand, gamma distributions [e.g. @McGill1965; @Wolfe2002] do model a random occurence of events (rate parameter) where only some fraction of these trigger an action (shape parameter). This could be consistent with the more frequent microsaccades happening randomly at a faster rate where only some of these would trigger a reset. In other words, the gamma distribution is not only a good fit to our reset data but it may reveal the processes underlying the resets.

The fit showed that reset times were well modeled by a broad gamma distribution, with resets occurring randomly in time (set by the rate and shape parameter) with an average rate of 1 reset every 1.3 _s_ over the 3 to 4 _s_ duration of the trials. However, we have to qualify our conclusions given the noisy nature of hand-drawn trajectories as a measure of the perceived trajectory and the reset points. To gauge the generality of our results, we compared them to those from a similar study that analyzed trajectory endpoints [pre-print: @Liu2021] to determine the presence and size of spontaneous resets. In their study, the internal speed of the Gabor texture was adjusted across conditions and participants so that the illusion always appeared locally downward (vertical, ignoring any resets) compared to its actual 45° direction. This generated a highly predictable offset of path length / $\sqrt{2}$ at the end of the trajectory if there were no spontaneous resets. Any measured offset less than the expected value (path length / $\sqrt{2}$ ) then revealed the presence and size of spontaneous resets. Liu et al. [-@Liu2021] used this technique to measure spontaneous resets for  3 different durations and 2 path lengths (see Fig. 6A).

The data from 10 participants showed that for a fixed path and illusion strength, longer durations produced smaller illusory offsets at the end (Fig 6B, light blue dots). This can be explained if spontaneous resets occurred randomly over time, so that longer durations would have more resets and a smaller final offset at the endpoint. Liu et al. [-@Liu2021] modeled this with a Poisson process having a constant chance of reset of ~0.7 / _s_ (Fig 6B, orange line). In comparison, in our experiment, the gamma distribution fit to the data of explicit resets (rate=$2.81$, shape=$3.74$), gave a ~0.75 / _s_ chance of a reset, a remarkaly similar estimate. That's why, we also used the parameters of our gamma distribution fit from data to generate the PSEs that should be found in Liu et al.'s experiment with no additional fitting (Fig 6B, dashed purple line). We see that the two findings, based on very different approaches, are mostly in agreement with each other.

```{r fig6, echo=FALSE, message=FALSE, fig.width=6, fig.height=4, fig.cap="\\label{fig:fig6}**Figure 6:** Spontaneous resets without hand trajectories from Liu et al. [-@Liu2021], re-analyzed here. **A:** The Gabor’s internal motion was adjusted for each condition and participant so that the perceived path was always locally straight down (yellow arrows). The 350 ms temporal gap then ensured a full reset [@Lisi2015] to the physical location when the Gabor continued on its path following the gap (blue arrow). Gaps were presented after 1, 2 or 3 seconds, and 2 or 4 _dva_ real paths and the offset measured between the pre- and post-gap paths (the vertical green lines) indicated the amount of illusion left at the end of  the pre-gap path (measured PSE). **B:** Percent reduction of the illusion (difference between measured offset and the offset expected with no intervening spontaneous resets  as a percentage of expected offset) as a function of path length and duration. The size of the illusion decreased over time for a fixed path length and local illusion angle. These data are consistent with resets occurring randomly over time at ~0.7 / _s_ modeled as a Poisson process."}
plotJumpPSEs(target='inline', individualDots=F, modelDots=F, gammaDots=F, modelLine=T, gammaLine=T, propExpect=T)
```
